Clinical prediction using similarity-based cohorts: A scoping review

### 2023-03-30 docs/Manuscript/Manuscript.Rmd
revise manuscript following discussions with collaborators

---
title: "Clinical prediction using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall
    email: 
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: Nova Southeastern University College of Osteopathic Medicine, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Department, Street, City, State, Zip
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background

@@

  Methods

@@

  Results

@@

  Conclusions

@@

@@

We modeled our literature review on PRISMA guidelines (PRISMA Group, 2021). Because we focus on methodologies rather than approaches, we deviated in some ways from these guidelines.

## PRISMA-S checklist

The PRISMA-S checklist comprises 16 items categorized into information sources and methods, search strategies, peer review, and managing records. We use the checklist to report the steps of our systematic review. Some items of the PRISMA-S checklist are intended for meta analysis and, therefore, did not apply to this study. We followed the checklist throughout the search process and explain the deviations from it in [Supplemental Table 1](https://docs.google.com/spreadsheets/d/1lKI5WWGIzo_hE9olkWWVCbLjCY9To5fiKPiefXUqgoU/edit).

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms. Through each platform, we searched those databases included by default.[^databases]

[^databases]: This means that we searched both MEDLINE and PubMed Central (PMC) through Pubmed (we did not exclude other databases, but our earliest included results post-date them) and that we searched the six indices of the Core Collection (the Science Citation Index Expanded, the Social Sciences Citation Index, the Arts & Humanities Citation Index, the Emerging Sources Citation Index, the Conference Proceedings Citation Index, and the Book Citation Index) as well as several regional databases through the Web of Science platform.

@@

We conducted all searches over 2021 July 19–22. We organized the results into a Zotero collection with one subcollection for each database. We then imported the pooled results to Covidence, which merged some duplicate entries.

@@

Following the determination of 25 sources aligning with our systematic review, we collected characteristics of included studies ([Supplemental Table 2](https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/edit)). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, evaluation/performance/comparison measures (& their values), name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
googlesheets4::read_sheet(
  ss = "1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo", sheet = 1L
) |>
  rename_with(snakecase::to_snake_case) |>
  arrange(date_of_publication) |> 
  transmute(
    Citation = paste0("@", zotero_key),
    Date = format(date_of_publication, "%Y %b %d"),
    Task = objective,
    Source = type_of_data,
    Type = range_of_data,
    `Cases` = number_of_cases_incidents,
    `Features` = number_of_predictors_features
  ) |>
  mutate(Task = ifelse(Task == "Recommendation", "Decision Support", Task)) |> 
  mutate(across(
    c(`Cases`, `Features`),
    sapply, FUN = paste0, collapse = "; "
  )) |> 
  mutate(across(
    everything(),
    # ~ stringr::str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) ->
  tab_synthesis
tab_synthesis |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis.",
      " 'Date' is the earliest the study is known to have been public."
    )
  )
```

@@

Another dominant characteristic of included studies is their remit. We categorized studies as "science" or "quality improvement" based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as "science", making the majority of studies from a "quality improvement" standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care. Lastly, we observed a commonality among 6 of the 25 included studies in having evaluated methods using leave-one-out cross-validation (LOOCV). The purpose of this measure is to estimate the overall performance of certain factors when used to make predictions, particularly utilized on smaller data sets, where models benefit greatly from larger training sets and additional model fitting is less costly.

@@

```{r tab:composite, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
googlesheets4::read_sheet(
  ss = "1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs", sheet = 1L
) |>
  rename_with(snakecase::to_snake_case) |>
  transmute(
    Citation = paste0("@", zotero_key),
    Elements = str_remove_all(approach_type_s, "\\[|\\]"),
    Terminology = term_s#,
    # Outcomes = outcome
  ) |> 
  mutate(Elements = str_replace_all(Elements, ",", ";")) |> 
  left_join(select(tab_synthesis, Citation, Date), by = "Citation") |> 
  arrange(Date) |> 
  select(-Date) |> 
  mutate(across(
    everything(),
    # ~ str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) |>
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:composite}",
      "Methodological elements of studies included in the synthesis. "
    )
  )
```

@@

## Limitations

@@

## Appendix

### Search strings

Here we reproduce the search strings and platform specifications used in our literature search.
We first finalized the **PubMed** search string below:

@@

`"patient similarity" OR`                     48
`"patient distance" OR`                         
`"patient connection" OR`                       
`"patient affinity" OR`                         
`"patient clustering"`                          
Refined search: `"cohort"`                      
------------------------------------------------

>>>>

---
title: "Clinical prediction using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: Nova Southeastern University College of Osteopathic Medicine, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: University at Florida, Smathers Libraries Academic Research & Consulting Services, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  **Background**

@@

  **Methods**

@@

  **Results**

@@

  **Conclusions**

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
googlesheets4::gs4_auth(email = "cornelioid@gmail.com")
Sys.sleep(10)
```

@@

## PRISMA checklist

We include a PRISMA checklist as Supplemental Table 1.
Because we focus on methodologies rather than approaches, we deviate in some ways from PRISMA guidelines. In particular, many items of the checklist, involving bias assessment and quantitative synthesis, are intended for meta-analyses and, therefore, did not apply to this study.

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms.

Through each platform, we searched those databases included by default.
This means that we searched both MEDLINE and PubMed Central (PMC) through Pubmed (we did not exclude other databases, but our earliest included results post-date them) and that we searched the six indices of the Core Collection (the Science Citation Index Expanded, the Social Sciences Citation Index, the Arts & Humanities Citation Index, the Emerging Sources Citation Index, the Conference Proceedings Citation Index, and the Book Citation Index) as well as several regional databases through the Web of Science platform.

@@

We conducted all searches over 2021 July 19–22. We tentatively excluded results from Google Scholar because they were missing abstracts, and later agreed to discard these results due to the lack of reproducibility of the search. We organized the remaining results into a Zotero collection with one subcollection for each database. We then imported the pooled results to Covidence, which merged some duplicate entries.

@@

Following the determination of 25 sources aligning with our review, we collected characteristics of included studies ([Supplemental Table 2](https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/edit)). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, evaluation/performance/comparison measures (& their values), name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
googlesheets4::read_sheet(
  ss = "1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo", sheet = 1L
) |>
  rename_with(snakecase::to_snake_case) |>
  arrange(date_of_publication) |> 
  transmute(
    Citation = paste0("@", zotero_key),
    Date = format(date_of_publication, "%Y %b %d"),
    Task = objective,
    Aim = ifelse(generalizable_knowledge == "Science", "Knowledge", "Practice"),
    Source = type_of_data,
    Type = range_of_data,
    `Cases` = number_of_cases_incidents,
    `Features` = number_of_predictors_features
  ) |>
  mutate(Task = ifelse(Task == "Recommendation", "Decision Support", Task)) |> 
  mutate(across(c(Cases, Features), sapply, FUN = format, big.mark = ",")) |>
  mutate(across(c(Cases, Features), sapply, FUN = paste0, collapse = "; ")) |>
  # mutate(across(c(Cases, Features), str_split, pattern = "; ")) |> 
  # mutate(Dimensions = purrr::map2_chr(
  #   Cases, Features,
  #   ~ str_c(.x, .y, sep = " × ", collapse = "; ")
  # )) |> 
  mutate(across(
    everything(),
    # ~ stringr::str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) ->
  tab_synthesis
tab_synthesis |> 
  select(-Date) |> 
  # prevent numeric list columns from dominating the horizontal space
  mutate(across(c(-Cases, -Features), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis,",
      " arranged by the earliest the study is known to have been public."
    )
  )
```

@@

Another dominant characteristic of included studies is their remit. We categorized studies as "knowledge" or "practice" based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as "knowledge", making the majority of studies from a "practice" standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care. Lastly, we observed a commonality among 6 of the 25 included studies in having evaluated methods using leave-one-out cross-validation (LOOCV). The purpose of this measure is to estimate the overall performance of certain factors when used to make predictions, particularly utilized on smaller data sets, where models benefit greatly from larger training sets and additional model fitting is less costly.

@@

```{r tab:composite, echo=FALSE, message=FALSE, warning=FALSE}
googlesheets4::read_sheet(
  ss = "1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs", sheet = 1L
) |>
  rename_with(snakecase::to_snake_case) |>
  transmute(
    Citation = paste0("@", zotero_key),
    Elements = str_remove_all(approach_type_s, "\\[|\\]"),
    Terminology = term_s#,
    # Outcomes = outcome
  ) |> 
  mutate(Elements = str_replace_all(Elements, ",", ";")) |> 
  left_join(select(tab_synthesis, Citation, Date), by = "Citation") |> 
  arrange(Date) |> 
  select(-Date) |> 
  mutate(across(
    everything(),
    # ~ str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) |>
  mutate(across(c(Citation), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:composite}",
      "Methodological elements of studies included in the synthesis. "
    )
  )
```

@@

@@

# Appendix

## Search strings

Here we reproduce the search strings and platform specifications used in our literature search.
We first finalized the **PubMed** search string below:[^pubmed-url]

[^pubmed-url]: <https://pubmed.ncbi.nlm.nih.gov/?term=(+"case-based+reasoning"+[All+Fields]+OR+"case-based+system"+[All+Fields]+)+OR+(+"individualized+modeling"+[All+Fields]+OR+"personalized+modeling"+[All+Fields]+OR+"customized+modeling"+[All+Fields]+)+OR+(+"individualized+cohort"+[All+Fields]+)+OR+(+(+"patient+similarity"+[All+Fields]+OR+"patient+distance"+[All+Fields]+OR+"patient+connection"+[All+Fields]+OR+"patient+affinity"+[All+Fields]+OR+"patient+clustering"+[All+Fields]+)+AND+(+"cohort+study"+[All+Fields]+)+)&sort=date>

@@

`"patient similarity" OR`                     48
`"patient distance" OR`                         
`"patient connection" OR`                       
`"patient affinity" OR`                         
`"patient clustering"`;                         
Refined search: `"cohort"`                      
------------------------------------------------
